### 2024-06-10 doc/sections/introduction.tex
[ tex ] Proof reading, and fixing some references

MAV and BV are \emph{Deep Inference} systems. Deep Inference~\cite{Guglielmi14:di} is generalisation of Gentzen's methodology for designing proof systems, which arose from Guglielmi's attempts to relate process algebra~\cite[CCS]{Milner80:CCS,Milner89:CC} to Classical Linear Logic~\cite[CLL]{Girard87:ll}.
The problem with such a relation is that, while the multiplicative connectives of Linear Logic capture parallel composition, no connective of Linear Logic captures \emph{sequential composition}.
Eventually, Guglielmi's attempts yielded BV, which extends Multiplicative Linear Logic~\cite[MLL]{Girard87:ll} with a self-dual non-commutative connective that captures sequential composition.
Such a connective was already present in another extension of Linear Logic, Pomset logic~\cite{Retore97:pomset}, where it arose from the study of coherence space semantics of Linear Logic~\cite[\C8]{GirardTL89:proofs}.
Recently, Nguyễn and Stra{\ss}burger~\cite{NguyenS22:bvisnotpl,NguyenS23:complexity} showed that, while BV is similar to Pomset logic, the two are not the same, as the theorems of BV form a proper subset of the theorems of Pomset logic.
Neither BV nor Pomset logic has a sequent calculus. Tiu~\cite{Tiu06:sisii} showed that sequent calculus cannot capture BV, and it is assumed that this result extends to Pomset logic.

\emph{Cut Elimination}, or \emph{admissibility of Cut}, is the fundamental property of Gentzen Sequent Calculi systems, which states that proofs using the Cut rule to introduce ``detours'' can be normalised to ones without. Crucial properties such as consistency and decidability follow from Cut-elimination. The Deep Inference analogue of Cut-elimination is the admissibility of the whole \emph{up} fragment of the calculus, which includes the Deep Inference form of Cut (which we describe in \Cref{sec:mav-syntax} below) as well as duals of most of the other rules of the calculus. Admissibility of the up fragment has the same metatheoretic benefits for Deep Inference systems as it does for Sequent Calculus ones.

@@

We present an alternative proof of cut-elimination via a semantic model. This proof avoids some of the intricacy of Horne's proof. We believe our proof to be more robust in the presence of extensions, due to our use of standard constructions such as Day monoids, order ideals, and the Chu construction. We demonstrate this by scaling down to plain BV and up to MAV with additive units.

@@

The phase semantic proof of cut-elimination does not easily extend to include the kind of self-dual connective present in BV and its extensions.
The phase space model derives duality by means of double negation with respect to the monoidal structure, which means that any connective has a derived dual. Attempts to extend the model with the non-commutative connective result in two distinct dual non-commutative connectives, reminiscent of the non-commutative tensor and par introduced by Slavnov~\cite{Slavnov19:scmll}.

@@

Our main contribution is the use of semantic techniques to derive the admissibility of identity expansion, cut, and the other \emph{co-} rules of MAV. To this end, we have developed a number of results on the semantics of BV and MAV:
\begin{enumerate}
      \item In \Cref{sec:mav-algebras}, we propose MAV-algebras as the algebraic counterpart of MAV. In short, an MAV-algebra is a $*$-autonomous partial order with meets, and another partially ordered monoidal structure that is \emph{duoidal} with respect to the $*$-autonomous structure.
      \item Normal proofs (our name for the ``up'' fragment) as we define them in the next section do not {\it a priori} support all the structure of an MAV-algebra, so we define the weaker notion of \emph{MAV-frame} (\Cref{defn:mav-frame}). These play the same role as Kripke structures do for modal and substructural logics. The first main technical contribution of the paper is in showing that every MAV-frame can be completed to an MAV-algebra.
      \item We then show that MAV-frames are strongly complete (in the terminology of Okada) for MAV by proving that the MAV-frame constructed from normal proofs can be used to deduce that all MAV-provable structures have normal proofs in \Cref{thm:cut-elim}. Completeness in the usual sense also follows in \Cref{thm:completeness}.
\end{enumerate}

@@

As far as we are aware, the semantics of BV and its extensions have not been considered before in the literature. The crucial part of the proof is that the standard Chu construction extends to the self-dual non-commutative connective of BV and MAV (\Cref{prop:chu-self-dual} and \Cref{prop:chu-monoid-duoidal}). Our development of the semantics of MAV and BV also opens the possibility of using these logics as sound and complete for reasoning about the MAV-frame structures we define.

>>>>

MAV and BV are \emph{Deep Inference} systems. Deep Inference~\cite{Guglielmi14:di} is generalisation of Gentzen's methodology for designing proof systems, which arose from Guglielmi's attempts to relate process algebra~\cite[CCS]{Milner80:CCS,Milner89:CC} to Classical Linear Logic~\cite[CLL]{Girard87:ll}.
The problem with such a relation is that, while the multiplicative connectives of Linear Logic capture parallel composition, no connective of Linear Logic captures \emph{sequential composition}.
Eventually, Guglielmi's attempts yielded BV, which extends Multiplicative Linear Logic~\cite[MLL]{Girard87:ll} with a self-dual non-commutative connective for sequential composition.
Such a connective was already present in another extension of Linear Logic, Pomset logic~\cite{Retore97:pomset}, where it arose from the study of coherence space semantics of Linear Logic~\cite[\C8]{GirardTL89:proofs}.
Recently, Nguyễn and Stra{\ss}burger~\cite{NguyenS22:bvisnotpl,NguyenS23:complexity} showed that, while BV is similar to Pomset logic, the two are not the same, as the theorems of BV form a proper subset of the theorems of Pomset logic.
Neither BV nor Pomset logic has a sequent calculus. Tiu~\cite{Tiu06:sisii} showed that no sequent calculus can capture BV, and it is assumed that this result extends to Pomset logic.

\emph{Cut-elimination}, or \emph{admissibility of Cut}, is the fundamental property of Gentzen Sequent Calculi systems, which states that proofs using the Cut rule to introduce ``detours'' can be normalised to ones without. Crucial properties such as consistency and decidability follow from Cut-elimination. The Deep Inference analogue of Cut-elimination is the admissibility of the whole \emph{up} fragment of the calculus, which includes the Deep Inference form of Cut (which we describe in \Cref{sec:mav-syntax} below) as well as duals of most of the other rules of the calculus. Admissibility of the \emph{up} fragment has the same metatheoretic benefits for Deep Inference systems as it does for Sequent Calculus ones.

@@

We present an alternative proof of cut-elimination via a semantic model. This proof avoids some of the intricacy of Horne's proof. We believe our proof to be more robust in the presence of extensions, due to our use of standard constructions such as Day monoids, order ideals, and the Chu construction. We demonstrate this by scaling down to plain BV and up to MAV with additive units, and also to BV with exponentials (Guglielmi and Stra{\ss}burger's System NEL \cite{Burger_2011,GuglielmiS11}).

@@

The phase semantic proof of cut-elimination does not easily extend to include the kind of self-dual connective present in BV and its extensions.
The phase space model derives duality by means of double negation with respect to the monoidal structure, which means that any connective has a derived dual. Attempts to extend the model with the non-commutative connective result in two distinct but dual non-commutative connectives, reminiscent of the non-commutative tensor and par introduced by Slavnov~\cite{Slavnov19:scmll}.

@@

Our main contribution is the use of semantic techniques to derive the admissibility of identity expansion, cut, and the other \emph{co-} rules of MAV. To this end, we have developed a number of results concerning the semantics of BV and MAV:
\begin{enumerate}
      \item In \Cref{sec:mav-algebras}, we propose MAV-algebras as the algebraic counterpart of MAV. In short, an MAV-algebra is a $*$-autonomous partial order with meets, with another partially ordered monoid structure that is \emph{duoidal} with respect to the $*$-autonomous structure.
      \item Normal proofs (our name for the ``up'' fragment) as we define them in the next section do not {\it prima facie} support all the structure of an MAV-algebra, so we define the weaker notion of \emph{MAV-frame} (\Cref{defn:mav-frame}). These play the same role as Kripke structures do for modal and substructural logics. The first main technical contribution of the paper is in showing that every MAV-frame can be completed to an MAV-algebra.
      \item We then show that MAV-frames are strongly complete (in the terminology of Okada) for MAV by proving that the MAV-frame constructed from normal proofs can be used to deduce that all MAV-provable structures have normal proofs in \Cref{thm:cut-elim}. Completeness in the usual sense also follows in \Cref{thm:completeness}.
\end{enumerate}

@@

As far as we are aware, the semantics of BV and its extensions have not been considered before in the literature. The crucial part of the proof is that the standard Chu construction extends to the self-dual non-commutative connective of BV and MAV (\Cref{prop:chu-self-dual} and \Cref{prop:chu-monoid-duoidal}). Our development of the semantics of MAV and BV also opens the possibility of using these logics as sound and complete systems for reasoning about the MAV-frame structures we define.

### 2024-06-10 doc/sections/mav-semantics.tex
[ tex ] fix typo in residuation
[ tex ] Proof reading, and fixing some references

To our knowledge, the semantics of MAV have not been previously
studied in the literature. The rules are an extension of MALL, so an
\emph{MAV-algebra} will partly be a $*$-autonomous partial order
(\Cref{defn:star-autonomous}) with meets and joins. The additional
structure for seq satisfies the conditions of a duoidal category
\cite[Definition 6.1]{Aguiar_2010} (\Cref{defn:duoidal}). We show that
MAV is sound for MAV-algebras in \Cref{thm:soundness}.

@@

The semantics of MAV is a collection of interacting monoids
on a partial order. We collect here the basic definitions and kinds of
interaction we will need.

@@

\begin{definition}\label[defn]{defn:residual}
  A commutative pomonoid $(\bullet, i)$ on a poset $(A, \leq)$ is
  \emph{residuated} if there is a function
  $\rightblackspoon : A \times A \to A$ such that $x \bullet y \leq z$
  iff $x \leq y \rightblackspoon y$.
\end{definition}

Linear logic adds a duality, or negation, to a commutative
pomonoid structure. Semantically, duality with commutativity is
captured in the definition of $*$-autonomous category is, due to Barr~\cite{Barr_1979}. For our purposes, we need the partial order
analogue, also called a CL algebra by Troelstra~\cite{Troelstra92:lll}.

@@

\begin{remark}
  The construction of the MAV-frame \NMAV does not use the $\vTens$
  and $\vPlus$ structure of MAV directly.  This structure is recovered
  by duality from the other connectives by the constructions in the
  rest of this section. This corresponds to the fact that the
  \textsc{Co-X} rules in MAV that we wish to show admissible are the
  ones that mention the $\otimes$ and $\oplus$ connectives, with the
  exception of \cref{rule:Switch}, which has a special role to play in
  \Cref{prop:embedding-sem} in mediating interaction.
\end{remark}

>>>>

To our knowledge, the semantics of MAV have not been previously
studied in the literature. The rules are an extension of MALL, so an
\emph{MAV-algebra} will partly be a $*$-autonomous partial order
(\Cref{defn:star-autonomous}) with meets and joins. The additional
structure for $\vSeq$ satisfies the conditions of a duoidal category
\cite[Definition 6.1]{Aguiar_2010} (\Cref{defn:duoidal}). We show that
MAV is sound for MAV-algebras in \Cref{thm:soundness}.

@@

The algebraic semantics of MAV is a collection of interacting monoids
on a partial order. We collect here the basic definitions and kinds of
interaction we will need.

@@

\begin{definition}\label[defn]{defn:residual}
  A commutative pomonoid $(\bullet, i)$ on a poset $(A, \leq)$ is
  \emph{residuated} if there is a function
  $\rightblackspoon : A \times A \to A$ such that $x \bullet y \leq z$
  iff $x \leq y \rightblackspoon z$.
\end{definition}

Linear logic adds a duality, or negation, to a commutative
pomonoid structure. Semantically, duality with commutativity is
captured in the definition of $*$-autonomous category, due to Barr~\cite{Barr_1979}. For our purposes, we need the partial order
analogue, also called a CL algebra by Troelstra~\cite{Troelstra92:lll}.

@@

\begin{remark}
  The construction of the MAV-frame \NMAV does not use the $\vTens$
  and $\vPlus$ structure of MAV directly.  This structure is recovered
  by duality from the other connectives by the constructions in the
  rest of this section. This corresponds to the fact that the
  \textsc{Co-X} rules in MAV that we wish to show admissible are the
  ones that mention the $\vTens$ and $\vPlus$ connectives, with the
  exception of \cref{rule:Switch}, which has a special role to play in
  \Cref{prop:embedding-sem} in mediating interaction.
\end{remark}

### 2024-06-10 doc/sections/mav-syntax.tex
[ tex ] fix typo in associtativity equations
Add nota bene
Fix nota bene
[ tex ] Proof reading, and fixing some references

In Deep Inference terminology, proofs operate on \emph{structures}, which simultaneously play the role of formulas and sequents in traditional Sequent Calculus systems. There are a number of different notations in the literature for the structures of BV and related systems. For familiarity's sake, we opt for a notation similar to normal Linear Logic, albeit extended with the self-dual non-commutative connective $\vSeq$.

@@

     & (\vP\vSeq\vQ)\vSeq\vP

@@

     & (\vP\vTens\vQ)\vTens\vP

@@

     & (\vP\vParr\vQ)\vParr\vP

@@

However, as BV's inference rules can work arbitrarily deep in the structure, the surrounding context of pars is unnecessary---and too restrictive. Hence, the cut for BV is rule (\ref{rule:CutBV}).

@@

Proof trees are a cumbersome presentation for BV's derivations---they are convenient for trees, but cumbersome in the absence of branching.
\emph{Rewriting systems}, on the other hand, are a well-known and convenient notation for such derivations.
Hence, in MAV, inference rules are presented as rewrite rules.

@@

@@

The inference rules come in dual pairs. For every rule $\vP\vInferFrom\vQ$, there is a dual rule $\vDual\vP\vInferFrom\vDual\vQ$.
The exception is \cref{rule:Switch}, which is self-dual, up to commutativity.
The \cref{rule:CoInteract}, \cref{rule:CoLeft}, and \cref{rule:CoRight} rules introduce of new structures going left-to-right.
Normal proofs, which we define below, avoid these \emph{synthetic} rules.

>>>>

In Deep Inference terminology, proofs operate on \emph{structures}, which simultaneously play the role of formulas and sequents in traditional Sequent Calculus systems. There are a number of different notations in the literature for the structures of BV and related systems. For familiarity's sake, we opt for a notation similar to the formulas of normal Linear Logic, albeit extended with the self-dual non-commutative connective $\vSeq$.

@@

     & (\vP\vSeq\vQ)\vSeq\vR

@@

     & (\vP\vTens\vQ)\vTens\vR

@@

     & (\vP\vParr\vQ)\vParr\vR

@@

However, as BV's inference rules can work arbitrarily deep in the structure, and the system contains the \cref{rule:Switch} rule, the surrounding context of $\vParr$s is unnecessary---and too restrictive. Hence, the cut for BV is rule (\ref{rule:CutBV}).

@@

Proof trees are a cumbersome presentation for BV's derivations---they are convenient for branching sequent proofs, but BV derivations are sequences of structures.
\emph{Rewriting systems}, on the other hand, are a well-known and convenient notation for such derivations.
Hence, in MAV, inference rules are presented as rewrite rules.

@@

\textbf{N.B.}\ $\vP\vInferFrom\vQ$ is an inference rule, \emph{not a sequent}. In sequent calculus notation, it is $\frac{\vQ}{\vP}$, \emph{not} $\vP\vdash\vQ$.

@@

The inference rules come in dual pairs. For every rule $\vP\vInferFrom\vQ$, there is a dual rule $\vDual\vQ\vInferFrom\vDual\vP$.
The exception is \cref{rule:Switch}, which is self-dual, up to commutativity.
The \cref{rule:CoInteract}, \cref{rule:CoLeft}, and \cref{rule:CoRight} rules introduce new structures going left-to-right.
Normal proofs, which we define below, avoid these \emph{synthetic} rules.
